---
title: "Lab2 - inferential statistics - confidence intervals"
author: "Isabel Sassoon"
date: "28/09/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Confidence Intervals in R

We will be showing you how to use R to compute the confidence intervals for a proportion and for a mean.

## Confidence interval for a proportion

A Market research agency interviews a random sample of 1000 people who live in London, and finds that 49% of people suggest they will vote Labour for the London Mayor. Find a 95% and a 90% confidence interval for the true proportion of voters who will vote for a Labour Mayor.

```{r}
prop.test(490, 1000, conf.level = 0.95)
```

95 percent confidence interval:
 0.46 0.52


## What about the 90% CI?


```{r}
prop.test(490, 1000, conf.level = 0.9)
```

The 90% confidence interval is

 0.46 0.52
 
In this case when apprximated to 2 decimal places the confidence intervals are almost identical! 

## Confidence interval for the mean

### Reading in from csv - use the skew data.csv

A bit more practice in reading data into R. Make sure you have the skewdata.csv file in the same location as your R project. In this case I have it stored in a subfolder called data.

```{r}
data<-read.csv("skewdata.csv")
```

## Visualise the data

```{r}
summary(data)
```

This data contains only one variable. We can see its mean is 30.969.


# What is the 95% confidence interval for this variable?

```{r}
t.test(data$values)
```

The 95 percent confidence interval: is 24.11 37.82
 
## We can do the same for the 90 percent confidence interval

```{r}
t.test(data$values, conf.level = 0.9)
```
90 percent confidence interval: 25.27 36.66

We can see that the confidence interval is smaller as we decrease our confidence level. 




