---
title: "Lab 2 - part 2"
output: html_document
date: "2022-09-30"
---

#Confidence Intervals

For this practice, we are going to use the "auto-mpg" dataset.

```{r}
df <- read.csv("auto-mpg.csv")
```

A brief summary:

```{r}
summary(df)
```

As seen, horsepower should also be numeric. Looking at the df, there are some values containing "?". We will replace these with Na:

```{r}
df[df == "?"] <- NA
```

Now we will obtain the statistics summary:

```{r}
summary(df$horsepower)
```
Something weird is still happening. We will need to convert "horsepower" from character to numeric

```{r}
df$horsepower <- as.numeric(df$horsepower)
```

Now lets check.

```{r}
summary(df$horsepower)
```

Now we have all the information correctly displayed. Now, we have to check which columns look normally distributed:

```{r}
names(df)
```
```{r}
library(ggplot2)
```


```{r}
ggplot(df)+ geom_bar(aes(x=mpg), fill="blue") +
  labs(title = "mpg") +
  scale_y_continuous(name="Frequency", breaks=c(0,5,10)) +
  theme_classic()
```

```{r}
ggplot(df)+ geom_bar(aes(x=cylinders), fill="blue") +
  labs(title = "cylinders") +
  scale_y_continuous(name="Frequency", breaks=c(0,5,10)) +
  theme_classic()
```

```{r}
ggplot(df)+ geom_bar(aes(x=displacement), fill="blue") +
  labs(title = "displacement") +
  scale_y_continuous(name="Frequency", breaks=c(0,5,10)) +
  theme_classic()
```
```{r}
ggplot(df)+ geom_bar(aes(x=horsepower), fill="blue") +
  labs(title = "horsepower") +
  scale_y_continuous(name="Frequency", breaks=c(0,5,10)) +
  theme_classic()
```

```{r}
ggplot(df)+ geom_bar(aes(x=weight), fill="blue") +
  labs(title = "weight") +
  scale_y_continuous(name="Frequency", breaks=c(0,5,10)) +
  theme_classic()
```
```{r}
ggplot(df)+ geom_bar(aes(x=acceleration), fill="blue") +
  labs(title = "acceleration") +
  scale_y_continuous(name="Frequency", breaks=c(0,5,10)) +
  theme_classic()
```

```{r}
ggplot(df)+ geom_bar(aes(x=model.year), fill="blue") +
  labs(title = "model year") +
  scale_y_continuous(name="Frequency", breaks=c(0,5,10)) +
  theme_classic()
```

As seen, the only variable apparently normally distributed is "acceleration". Let's obtain it's 95% CI:

```{r}
t.test(df$acceleration)
```
Being (15.23, 15.84)

Now, what proportion of cars are form the 80s?

```{r}
length(which(df$model.year >= 80))
```
```{r}
89/length(df$model.year)
```
There is an easier method to compute this:

```{r}
mean(df$model.year >= 80)
```

Thus, the proportion of cars form the 80s is 0.22. Now, we are asked to compute the 90% CI for this proportion:

```{r}
prop.test(89, length(df$model.year), conf.level = 0.9)
```

Although, we know this answer is NOT correct, as the variable "model year" is not normally distributed. For this, we would have to apply a bootstrap:

```{r}
library(boot)
```

```{r}
set.seed(1)
func <- function(data,i){
  d <- data[i,]
  c(
    mean(d[,7])
  )
}
```

```{r}
boot1 <- boot(df,func,R=1000)
boot1
```
```{r}
boot.ci(boot1,index=1)
```
In this process, we have calculated the 95% CI for the mean of the "model.year" variable. Now, we will calculate the 90% CI for the proportion of cars produced later than the 80s:

```{r}
set.seed(2)
func1 <- function(data,i){
  d <- data[i,]
  c(
    mean(d[,7] >= 80)
  )
}
```

```{r}
boot2 <- boot(df,func1,R=1000)
boot2
```
```{r}
boot.ci(boot2,index=1, conf = 0.9)
```
So the 90% CI for the proportion of cars made past the 80s is (0.1887,  0.2558).

#Distributions

Assume that among diabetics the fasting blood level of glucose is approximately normally
distributed with a mean of 105 mg per 100 ml and a standard deviation of 9mg per 100ml.

To visualize it:

```{r}
curve(dnorm(x, 105, 9), from=92, to=118)
```
Now, to find the proportion of diabetics that have levels between 90 and 125 mg per 100ml, we can do the following:

```{r}
level125 <- pnorm(125, 105, 9)
level125
```
```{r}
level90 <- pnorm(90, 105, 9)
level90
```
```{r}
answer <- level125-level90
answer
```
Being that the final answer.

Now we want to know what level cuts off the lower 10% of diabetics:

```{r}
qnorm(0.1,mean=105,sd=9)
```
Therefore, 10% of diabetics will have a level of glucose of 93.46 mg or lower.
