---
title: "QDA Lab 1 Part 2 A solution"
author: "Dr Isabel Sasssoon"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

This is a solution to the independent practice in the first lab  . In R there are many ways of obtaining the same information.

This data set is available in the *{ggplot}* library that was already loaded earlier in this session.


If you have not loaded ggplot2 then the code is below:  

```{r loading-libs}
library(ggplot2)
```

(a) Run the code below to see the description of the diamond data and its contents.

```{r}
help("diamonds")
```

Now the data can be accessed, using head(data) returns the first few rows of data.

```{r head-data}
head(diamonds)
```

(b) Check that the data set *diamonds* does indeed contain the columns as in the help

```{r var-names}
names(diamonds)
```

or to have more details on the columns:

```{r}
str(diamonds)
```

(c) The help description of the data set has the ranges for the variables in this data set. Check that this is the case in the data. 
Are there any variables out of range?

The ranges for the numerical variables can be obtained using the summary function. There are other ways.

To run summary on one variable only use the \$ to specify the column.

```{r var-summary}
summary(diamonds$price)
```

The code above can be repeated for each of the variables in turn or use summary on all the data frame.

```{r}
summary(diamonds)
```

It seems like there is an inconsistency in column z, the table shows a range of 0-3.18 but the actual data tells a different story. *Always check what is in the data*

(d) How many levels (or different categories) are in the column labelled *cut*? 

The levels of a categorical variable can be obtained using a table function.

```{r}
table(diamonds$cut)
```

The variable cut is categorical and has 5 levels or values.

(e) Generate a numerical summary for each variable in the data set - take care to use the appropriate technique for each different variable type.

The numerical summary can be done as in (c), but in general for numerical attributes a summary works well, a table for categorical variables.

```{r}
summary(diamonds)
```

(f) For each variable generate a plot, use different options and make sure the plots are labelled, and suitable to the type of variable.

For example for a numerical attribute we would look at a histogram

Firstly we can use the graph function in base R (this does not need a separate library)

```{r}
hist(diamonds$price, main=paste("Histogram of Price"), xlab="Price")
```

(optional) another way to plot this can be achieved using ggplot library:

```{r}
ggplot(diamonds, aes(x=price)) + geom_histogram(binwidth = 10) +
  theme_classic() + ggtitle("Histogram of the Diamond Price")
```

The same code can be used for the remaining continuous variables.

For the categorical variables we can use this shortcut that uses a table function to then make a plot.

```{r}
barplot(table(diamonds$cut), xlab = "Cut", ylab = "Frequency", main="Bar chart of the Frequency of each Diamond cut")
```

(Optional) If we want to use the GGplot library then the following ggplot syntax can be used:

```{r}
ggplot(diamonds, aes(cut)) + geom_bar() + ggtitle("Frequency of different cuts") +theme_classic()
```

Other options include pie charts.

(g) Explore the graphing of one categorical against one of the continuous variables in this data.

This is a good opportunity to use a box plot.

```{r}
boxplot(diamonds$price~diamonds$cut, xlab="Cut", ylab="Price")
```


```{r}
ggplot(diamonds, aes(x=cut, y=price)) + geom_boxplot() +
  theme_classic() + ggtitle("Price vs Cut for Diamonds")
```

You can try the same changing the variable on the y axis.
