---
title: "lab-6-independent-practice-a-solution"
author: "Isabel Sassoon"
date: "2022-10-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Lab 6 - Independent Practice - A solution

This is a solution to the worksheet for the QDA lab 6 independent practice. 
Different maximal models can be selected and as such the resulting minimal adequate model will differ. The important part is to justify the approach and the steps taken.

## Load required libraries

```{r}
library(ggplot2)
```


## 1. Load the *crime-analysis-data.csv* data into an R.

```{r}
crime<-read.csv("crime-analysis-data.csv")
```


## 2. Explore the data. Confirm the variables that are categorical and numerical/continuous and that R has read them in appropriately.

Starting with a summary of the contents:

```{r}
summary(crime)
```

The variable HighYouthUnemploy should be a categorical variable (actually binary as it only has two levels). R has read it in as numerical so this can be fixed by making it into a Factor.

```{r}
crime$HighYouthUnemploy<-as.factor(crime$HighYouthUnemploy)
```

Lets look at the distribution of the variables:

```{r}
ggplot(data = crime, aes(x=Youth)) + geom_histogram(bins = 20) + theme_classic()+ ggtitle("Youth Distribution")
ggplot(data = crime, aes(x=Education)) + geom_histogram(bins = 20) + theme_classic()+ ggtitle("Education Distribution")
ggplot(data = crime, aes(x=LabourForce)) + geom_histogram(bins = 20) + theme_classic() + ggtitle("Labour Force Distribution")
ggplot(data = crime, aes(x=ExpenditureYear0)) + geom_histogram(bins = 20) + theme_classic() + ggtitle("Expenditure Distribution")
ggplot(data = crime, aes(x=YouthUnemployment)) + geom_histogram(bins = 20) + theme_classic()+ ggtitle("Youth Unemployment Distribution")
ggplot(data = crime, aes(x=MatureUnemployment)) + geom_histogram(bins = 20) + theme_classic() + ggtitle("Mature Unemployment Distribution")
ggplot(data = crime, aes(x=Wage)) + geom_histogram(bins = 20) + theme_classic() + ggtitle("Wage Distribution")
ggplot(data = crime, aes(x=StateSize)) + geom_histogram(bins = 20) + theme_classic() + ggtitle("State Size Distribution")

```

The distribution for ExpenditureYear0 seems skewed to the left, other variables look generally symmetric. This does not warrant any transformations at this stage.


```{r}
ggplot(data = crime, aes(x=Youth, y=CrimeRate)) + geom_point() + theme_classic() + ggtitle("Youth vs Crime Rate")
ggplot(data = crime, aes(x=Education, y=CrimeRate)) + geom_point() + theme_classic()+ ggtitle("Education vs Crime Rate")
ggplot(data = crime, aes(x=LabourForce, y=CrimeRate)) + geom_point()  + theme_classic()+ ggtitle("Labour Force vs Crime Rate")
ggplot(data = crime, aes(x=ExpenditureYear0, y=CrimeRate)) + geom_point()  + theme_classic() + ggtitle("Expenditure vs Crime Rate")
ggplot(data = crime, aes(x=YouthUnemployment, y=CrimeRate)) + geom_point()  + theme_classic()+ ggtitle("Youth Unemployment vs Crime Rate")
ggplot(data = crime, aes(x=MatureUnemployment, y=CrimeRate)) + geom_point()  + theme_classic() + ggtitle("Mature Unemployment vs Crime Rate")
ggplot(data = crime, aes(x=Wage, y=CrimeRate)) + geom_point()  + theme_classic()+ ggtitle("Wage vs Crime Rate")
ggplot(data = crime, aes(x=StateSize, y=CrimeRate)) + geom_point() + theme_classic()+ ggtitle("State Size vs Crime Rate")

```

The collection of scatter plots do not show that any of the variables is clearly linear, but some show a linear trend.

## 3. Focusing only on the continuous explanatory variables - check their correlations with the CrimeRate. 
We want to do this only for the continuous variables, so can look to remove the column that is binary from this plot. (This is done so that the pairs plot is legible and that we can run a corr function on the resulting dataframe)

```{r}
crime.cont<-subset(crime, select=c("CrimeRate", "Youth", "Education", "ExpenditureYear0", "LabourForce",  "YouthUnemployment", "MatureUnemployment",
                                   "Wage", "StateSize") )
pairs(crime.cont)
```

```{r}
cor(crime.cont)
```

There do not seem to be any obvious multi collinearity (highly correlated explanatory variables) and a few of the plots above point to potential for a linear relationships, therefore at this stage I am not going to explore any transformations.


## 4. Using the continuous explanatory variables decide on a maximal model for CrimeRate and run it.

```{r}
crime.lm<-lm(crime$CrimeRate~crime$Youth+crime$Education+crime$ExpenditureYear0+crime$MatureUnemployment + crime$LabourForce+crime$YouthUnemployment+crime$StateSize + crime$Wage)

summary(crime.lm)
```

(NOTE: it is possible to start with a model that has interactions, all interactions could be used or a Tree approach can help understand if the relationship between an explanatory variable and the target variable is different based on the value (or range) of the explanatory variable - page 199 Crawley)

## 5. Use a model selection approach to achieve a minimal adequate model

Initially I will use all the continuous variables as explanatory variables as my starting maximal model. Then use the step function:

```{r}
step(crime.lm)
```

The step function has ended with this minimal adequate model:

crime$CrimeRate ~ crime$Youth + crime$ExpenditureYear0 + crime$MatureUnemployment + 
    crime$LabourForce
    

## 6. Once you have the minimal adequate model (from the step function), explain its findings and test its residuals

```{r}
mam.lm<-lm(formula = crime$CrimeRate ~ crime$Youth + crime$ExpenditureYear0 + 
    crime$MatureUnemployment + crime$LabourForce)

summary(mam.lm)
```
From the model summary for this minimal adequate model (mam.lm) we can see that crime rate is dependent on the variables Youth, ExpenditureYear0   and LabourForce. In all these cases an increase in their value increases the crime rate estimate.

$$CrimeRate= -230.01795 +1.02436\times Youth+ 0.77626 \times ExpenditureYear0 + 0.80540 \times MatureUnemployment +   0.17379 \times LabourForce $$

This model has acceptable goodness of fit, all the coefficients are significant (so there is no need to simplify further), $r^2$ is acceptable and the F statistic is significant.

Next the residuals should be scrutinized:

```{r}
plot(mam.lm)
```
In this case the residuals look ok, the variance is quite steady in the first plot - considering the data size.
QQ plot also looks aligned, with a few points standing out as outliers. We could explore these in more detail but as the model is good there may be no need to.


## 7. OPTIONAL -  model the relationship between the crime rate and the explanatory variables (including the ones that are not continuous).

Instead of only including the continuous variables in the model we can now add them all in, including the factor variable. 

```{r}
model.all.lm<-lm(crime$CrimeRate~crime$Youth+crime$Education+crime$ExpenditureYear0
                 +crime$MatureUnemployment + crime$LabourForce+crime$YouthUnemployment+crime$StateSize + crime$Wage+ crime$HighYouthUnemploy)

summary(model.all.lm)
```

The $r^2 $is higher than for the model without the factors but lets see what a step process would achieve in terms of simplifying the model:

```{r}
step(model.all.lm)
```

The binary (factor) variable we added as part of the explanatory variables does not add much and this is confirmed as the step process proposes a model that does not include it as an explanatory variable.


```{r}
ggplot(crime, aes(x=HighYouthUnemploy, y=CrimeRate)) + geom_boxplot() + ggtitle("Crime rate by High Youth Unemployment indicator")+ theme_classic()
```

Therefore the model from section 6 is probably still the most suitable one to use to answer the question asked "what variables affect the crime rate and how?".


## 8. OPTIONAL - If the average education time in the population is 14 years. Compute the mean education time in this sample of 48 rows of data and test the hypothesis that the education time is significantly lower than the population education time.


```{r}
education.mean<-mean(crime$Education)
education.mean
```

We will be testing the following hypotheses: $H_0: \mu=14$ and $H_1: \mu < 14$.

```{r}
t.test(crime$Education, mu=14, alternative = "less" )
```

This is a very small p-value, so we have evidence to reject the null hypothesis in this case. The mean education for this sample of data is significantly lower than that for the population.


***

### OPTIONAL - If you want to use a Regression Tree to check for interactions:

```{r}
library(tree)
crime.tree<-tree(crime$CrimeRate~crime$Youth+crime$Education+crime$ExpenditureYear0+crime$MatureUnemployment + crime$LabourForce+crime$YouthUnemployment+crime$StateSize + crime$Wage)
plot(crime.tree)
text(crime.tree)
```

From this tree there are no "contradictions" in direction of the estimate. (see page 199 of Crawley for an example), therefore this does not point to the need for interactions in this case.